Late Pleistocene stone artefact technology at Mau A, Yen Bai, Northern Vietnam

### 2022-03-25 analysis/paper/paper.Rmd
Removal of a few excess spaces
editing the text, final WI21 meeting

The late Pleistocene of Southeast Asia is characterized by flaked stone tools which rarely have strongly patterned forms (Mijares 2008; Borel et al. 2013; Borel et al. 2017). This lack of clear typo-technological categories for flakes makes understanding the cultural dynamics of period difficult. Previous papers have employed multivariate analysis of flake outlines in relation to other variables, such as the traces of use wear and retouching (Borel et al. 2013; Borel et al. 2017). Elliptical Fourier analysis was used to approximate shape outlines as sums of sine and cosine waves, with the coefficients from these sums being passed through Principal Component Analysis (Borel et al. 2013; Borel et al. 2017). This allowed the outlines of stone tools to be represented as points on a two-dimensional plane, facilitating comparisons between tool morphology and other variables, such as use wear. The use of outlines and not the raw linear measurements commonly employed in archaeology was intentional. Borel et al. (2017) specifically highlighted the dangers of using raw linear measurements, such as length and width, to describe tool morphology, as raw linear measurements can often fail to cover the full variation in shape often seen in stone tools.

@@

Flake shape variation generally shows limited correspondence with other factors. Figure \@ref(fig:excavation-unit-biplot)B shows very little variation over time, with lots of overlap between groups in the biplot. Units with fewer flakes, such as 13, show lower shape variation. This generally confirms our conclusion from the radiocarbon ages that the deposit accumulated relatively quickly, and the assemblage represents knapping events occurring at a similar time with little, if any perceptible change over time. By contrast, a more definitive difference between groups in seen in Figure \@ref(fig:excavation-unit-biplot)C, where there is a high degree of shape variation in the lightest flakes. This reflects the production of different types and shapes of cores as the flakes become smaller and thus lighter. This is further supported by Figure \@ref(fig:excavation-unit-biplot)A where tertiary flakes with little dorsal cortex show the highest shape variation. This supports our interpretation of the previous result, with flakes from later in the reduction process showing greater shape variation as different types of cores were produced at the site. 
The lack of distinct morphological groupings and the increased variability of shape in the smaller, tertiary flakes is consistent with previous studies. Other studies which investigated the morphology of flake outlines from South East Asia concluded that there is no significant relationship between flake outline and flake use (Borel et al., 2013; Borel et al., 2017). While our analysis not rely upon outlines, both outlines and landmarks are ultimately methods for measuring the morphology of flaked stone tools, and as such, the same models which explain outline variability can be cautiously applied to our analysis. The large amount of variability in flake shape, especially in the tertiary flakes, suggests that the knappers were not looking to create a specific shape of flake, with the amount of variability increasing as flakes are struck off of increasingly differed cores. This behavior is consistent with processes of expedient knapping, where a large number of flakes are created in a relatively free-form manner and useful flakes are picked up and used after knapping is finished (Holdaway & Douglass, 2012; McCall, 2012). Processes of expedient knapping suggest that stone tools were of secondary importance in the material culture of the knappers (Sillitoe & Hardy, 2003). Elsewhere it has been suggested stone tools were primarily used to create bamboo tools in South East Asia (Mijares, 2001). A reliance upon bamboo tools, with stone flakes merely being an intermediary to produce bamboo tools, would explain the great variability in the lithic assemblages at Mau A.

Looking at \@ref(fig:contribution-plots), we can see that certain landmarks contributed to each principal component. For the first principal component, the distal tip, the top right, and the top left landmarks are the top three contributors. We interpret this as indicating that the first principal component is primarily influenced by artefact length and platform breadth. The second principal component primarily consists of the bottom left, bottom right, top left, and top right landmarks; one can interpret this as the second principal component taking after overall width.

@@

>>>>

<!-- also make sure to provide a general context of paleolithic southeast asia. See previously suggested papers and look at context papers cited by Borel et al.  -->

Stone artefact manufacture in Mainland Southeast Asia (MSEA) may have began as early as 800 ka ago (Forestier et al., 2022), but many of these claims are contested (Marwick et al. 2022) and it is not until the Late Pleistocene that an indisputably strong signal of artefact production is evidence. The oldest skeletal remains of anatomically modern humans date to roughly 50-70 ka ago (Demeter et al., 2017; Shackelford et al., 2018). By the Last Glacial Maximum at MIS 2 many areas of the region were occupied by stone artefact using forager groups. In Northern Vietnam this time was characterized by sporadic human occupation of caves, localized differences in climate (with shifting cover of rain forest between the moist highlands and the variable lowlands) and the accumulation of mollusk shell middens (McAdams et al., 2020; 2022). Throughout the Late Pleistocene humans in MSEA would repeatedly reuse caves, accumulating numerous layers of discarded shells, ash, and lithics over multiple periods of occupation, with intervening periods of abandonment and occupation by faunal species such as bats (McAdams et al., 2020).

Lithic technologies in the wider Southeast Asian region during this time are characterized by flaked stone tools which rarely have strongly patterned forms (Mijares 2008; Borel et al. 2013; Borel et al. 2017). This lack of clear typo-technological categories for flakes makes understanding the cultural dynamics of period difficult, relative to regions with more visually distinctive technologies such as northern Europe and parts of Africa and South America. In their attempt to ... Borel et al. (2013; 2017) have employed multivariate analysis of flake outlines in relation to other variables, such as the traces of use wear and retouching. They used elliptical Fourier analysis to approximate artefact shape outlines as sums of sine and cosine waves, with the coefficients from these sums being passed through Principal Component Analysis (Borel et al. 2013; 2017). This allowed the outlines of stone tools to be reduced and projected onto a two-dimensional plane, facilitating comparisons between tool morphology and other variables, such as use wear. Borel et al. (2017) choose to use of outlines rather than traditional linear measurements because the claims that often fail to cover the full variation in shape often seen in stone tools.

@@

Flake shape variation generally shows limited correspondence with other factors. Figure \@ref(fig:excavation-unit-biplot)B shows very little variation over time, with lots of overlap between groups in the biplot. Units with fewer flakes, such as 13, show lower shape variation. This generally confirms our conclusion from the radiocarbon ages that the deposit accumulated relatively quickly, and the assemblage represents knapping events occurring at a similar time with little, if any perceptible change over time. By contrast, a more definitive difference between groups in seen in Figure \@ref(fig:excavation-unit-biplot)C, where there is a high degree of shape variation in the lightest flakes. This reflects the production of different types and shapes of cores as the flakes become smaller and thus lighter. This is further supported by Figure \@ref(fig:excavation-unit-biplot)A where tertiary flakes with little dorsal cortex show the highest shape variation. This supports our interpretation of the previous result, with flakes from later in the reduction process showing greater shape variation as different types of cores were produced at the site.

The lack of distinct morphological groupings and the increased variability of shape in the smaller, tertiary flakes is consistent with previous studies. Other studies which investigated the morphology of flake outlines from South East Asia concluded that there is no significant relationship between flake outline and flake use (Borel et al., 2013; Borel et al., 2017). While our analysis not rely upon outlines, both outlines and landmarks are ultimately methods for measuring the morphology of flaked stone tools, and as such, the same models which explain outline variability can be cautiously applied to our analysis. The large amount of variability in flake shape, especially in the tertiary flakes, suggests that the knappers were not looking to create a specific shape of flake, with the amount of variability increasing as flakes are struck off of increasingly differed cores. This behavior is consistent with processes of expedient knapping, where a large number of flakes are created in a relatively free-form manner and useful flakes are picked up and used after knapping is finished (Holdaway & Douglass, 2012; McCall, 2012). Processes of expedient knapping suggest that stone tools were of secondary importance in the material culture of the knappers (Sillitoe & Hardy, 2003). Elsewhere it has been suggested stone tools were primarily used to create bamboo tools in South East Asia (Mijares, 2001). A reliance upon bamboo tools, with stone flakes merely being an intermediary to produce bamboo tools, would explain the great variability in the lithic assemblages at Mau A.

In Figure \@ref(fig:contribution-plots), we can see that the distal tip, the top right, and the top left landmarks are the top three contributors to each principal component. We interpret this as indicating that the first principal component is primarily influenced by artefact length and platform width. The second principal component primarily consists of the bottom left, bottom right, top left, and top right landmarks; we interpret this as the second principal component representing surface area of the artefact.

@@

This work demonstrates the applicability of our new method in identifying the shape changes that accompany lithic reduction within a paleolithic MSEA assemblage. Our method for converting linear measurements into shape landmarks demonstrates a sensitivity to morphological diversity that opens the door for further applications. Our finding support the conclusions of Borel et al. (2017) that linear measurements rarely capture the full breadth of variation seen in stone tools. Our approach of taking linear measurements and converting them into landmarks gives more options for examining the sensitivity of shape variation to a variety of factors compared to the use of linear measurements alone. While the generated landmarks that we use here might not be as sensitive as full artefact outlines, linear measurements of stone tools are simple to collect with basic equipment. As a result, there are many opportunities to apply our method to existing linear measurement data on legacy assemblage.
